Fractional Types

### 2019-11-23 fracGC/p.tex
Pass over sec 4 to 6

We will call this language $\Pi/D$ for the fractional extension of
$\Pi$ with a dynamic check. In addition to being more expressive than
the language in the previous subsection, $\Pi/D$ has the advantage of
replacing the cumbersome cardinality-based evidence that the type is
not empty by the more intuitive evidence of choosing a value of the
type. The simplications are evident in the Agda formalization:

@@

\item Operational: The other reason our types can be justifiably
  called fractional is information theoretic. As mentioned in the
  introduction the number of bits needed to store information about a
  space with $n$ distinct values is $\log{n}$. If the number of bits
  can be thought of as a debt (to be reclaimed) then a negative number
  of bits corresponds to a space with $\oneover{n}$ distinct
  values. Indeed we interpret these values as gc processes each
  specialized to collect space for one of the $n$ distinct values. In
  this section, we make this intuition precise by presenting a
  (fully-formalized in Agda) abstract machine that manages an explicit
  heap.
\end{itemize}

@@

To make the space usage explicit in every step of evaluation, we
define an abstract machine whose states are pairs of a code register
and a memory register. In a low-level implementation the code register
would contain a pointer to the current instruction and the memory
register would contain a bit-level runtime representation of values
with one bit for boolean values, etc. In the Agda formalization, we
include the current combinator in the code register, and represent the
memory register using an index into an enumeration. Thus a memory
register containing a boolean value would be represented as
\AgdaFunction{[𝔽 ,  𝕋 ][0]} or \AgdaFunction{[𝔽 ,  𝕋 ][1]} depending on whether the
values is \AgdaInductiveConstructor{false} or \AgdaInductiveConstructor{true}.
The prepended vector
contains all possible values of the type and the number in between [\ldots ]
is an index into the vector, representing a finite number that
is guaranteed to be in bounds. In a low-level runtime representation
only a binary encoding of this finite type would need to be stored, but in
order to just prove that space is preserved, we can just inspect the
size of the prepended list. Putting it all together, machine states and
small-step reductions are formalized as follows:

@@

%%%%%
\subsection{Lifting Evaluation to Type System}

@@

The combinators in the lifted language $\Pi/\bullet$ consist of all the combinators
in the core $\Pi$ language but now operating on pointed types. The mediation
between general pointed types and singleton types is done via
\Acon{return} and \Acon{extract} which as we prove below form a
dual monad/comonad pair. The types for \Acon{$\eta$} and
\Acon{$\epsilon$} are now specialized to guarantee safety of
de-allocation as follows. When applying \Acon{$\eta$} at a pointed type, the
current witness value is put in focus in a singleton type and a gc
process for that particular singleton type is created. To apply this
process using \Acon{$\epsilon$} the very same singleton value must be
the current one.

@@

The mediation between pointed and singleton types is based a dual
monad/comonad pair from which many structural properties can be
derived.

\begin{proposition}
  $\llparenthesis \cdot \rrparenthesis$ is both an idempotent strong
  monad and an idempotent costrong comonad over pointed types.
\end{proposition}
\begin{proof}
  The main insight needed is to define the functor \Afun{∙Singᵤ}, the
  tensor/cotensor, and the join/cojoin (duplicate):
\PIPFCombderive{}
\end{proof}

@@

Thus, to summarize if a combinator expects a singleton type, then it
would only typecheck in the lifted language, if it is given the unique
value it expects. A particularly intriguing instance of that
situation is the following program:

@@

The program takes a value of type
\Acon{∙𝟙/}~(\Acon{∙𝟙/}~\Avar{A}). This would be a gc process
specialized to collect another gc process! At runtime type there would
be no information other than the functions that ignore their argument
but the type system provides enough guarantees to ensure that this
process is well-defined, is reversible, and is safe. The following
trace shows the actual execution for confirmation:

@@

\paragraph*{Space Transformers.} A pair of a value and a gc process
cancel each other when the gc process is expecting this particular
value. In more general cases, the pair can be viewed as a space
transformer that represents an implicit function that converts the value to the
one expected by the gc process. Such space transformers can be
composed and applied

@@

The feedback loop in the diagram is realized using ancillae allocation
and de-allocation. The safety condition on de-allocation ensures that
the allocated value (now in the feedback loop) matches the
de-allocated value. Thus, the entire traced component is treated as an
ancillae value used internally used to perform the larger program but
leaving no memory footprint.

>>>>

We will call this language $\Pi/D$ for the fractional extension of
$\Pi$ with a dynamic check. In addition to being more expressive than
the language in the previous subsection, $\Pi/D$ has the advantage of
replacing the cumbersome cardinality-based evidence that the type is
not empty by the more intuitive evidence of choosing a value of the
type. The simplifications are evident in the Agda formalization:

@@

\item Operational: The other reason our types can be justifiably
  called fractional is information theoretic. As mentioned in the
  introduction, the number of bits needed to store information about a
  space with $n$ distinct values is $\log{n}$. If the number of bits
  can be thought of as a debt (to be reclaimed) then a negative number
  of bits corresponds to a space with $\oneover{n}$ distinct
  values. Indeed we interpret these values as gc processes each
  specialized to collect space for one of the $n$ distinct values. In
  this section, we make this intuition precise by presenting a
  (fully-formalized in Agda) abstract machine that manages an explicit
  heap.
\end{itemize}

@@

To make the space usage explicit in every step of evaluation, we
define an abstract machine whose states are pairs of a code register
and a memory register. In a low-level implementation the code register
would contain a pointer to the current instruction and the memory
register would contain a bit-level runtime representation of values
with one bit for boolean values, etc. In the Agda formalization, we
include the current combinator in the code register, and represent the
memory register using an index into an enumeration. Thus a memory
register containing a boolean value would be represented as
\AgdaFunction{[𝔽 ,  𝕋 ][0]} or \AgdaFunction{[𝔽 ,  𝕋 ][1]} depending on whether the
value is \AgdaInductiveConstructor{false} or \AgdaInductiveConstructor{true}.
The prepended vector
contains all possible values of the type and the number in between [\ldots ]
is an index into the vector, representing a finite number that
is guaranteed to be in bounds. In a low-level runtime representation
only a binary encoding of this finite type would need to be stored, but in
order to just prove that space is preserved, we can just inspect the
size of the prepended list. Putting it all together, machine states and
small-step reductions are formalized as follows:

@@

%%%%%
\subsection{Lifting Evaluation to the Type System}

@@

The combinators in the lifted language $\Pi/\bullet$ consist of all the combinators
in the core $\Pi$ language but now operating on pointed types. The mediation
between general pointed types and singleton types is done via
\Acon{return} and \Acon{extract}, which as we prove below form a
dual monad/comonad pair. The types for \Acon{$\eta$} and
\Acon{$\epsilon$} are now specialized to guarantee safety of
de-allocation as follows. When applying \Acon{$\eta$} at a pointed type, the
current witness value is put in focus in a singleton type and a gc
process for that particular singleton type is created. To apply this
process using \Acon{$\epsilon$} the very same singleton value must be
the current one.

@@

The mediation between pointed and singleton types is based on a dual
monad/comonad pair from which many structural properties can be
derived.

\begin{proposition}
  $\llparenthesis \cdot \rrparenthesis$ is both an idempotent strong
  monad and an idempotent costrong comonad over pointed types.
\end{proposition}
\begin{proof}
  The main insight needed is to define the functor \Afun{∙Singᵤ}, the
  \Afun{tensor}/\Afun{cotensor}, and the \Afun{join}/\Afun{cojoin}
  (\Afun{duplicate}):
\PIPFCombderive{}
\end{proof}

@@

Thus, to summarize, if a combinator expects a singleton type, then it
would only typecheck in the lifted language, if it is given the unique
value it expects. A particularly intriguing instance of that
situation is the following program:

@@

The program takes a value of type
\Acon{∙𝟙/}~(\Acon{∙𝟙/}~\Avar{A}). This would be a gc process
specialized to collect another gc process! At runtime, there would
be no information other than the functions that ignore their argument
but the type system provides enough guarantees to ensure that this
process is well-defined, is reversible, and is safe. The following
trace shows the actual execution for confirmation:

@@

\paragraph*{Space Transformers.} A pair of a value and a gc process
cancel each other when the gc process is expecting this particular
value. In more general cases, the pair can be viewed as a space
transformer that represents an implicit function that converts the value to the
one expected by the gc process. Such space transformers can be
composed and applied.

@@

The feedback loop in the diagram is realized using ancillae allocation
and de-allocation. The safety condition on de-allocation ensures that
the allocated value (now in the feedback loop) matches the
de-allocated value. Thus, the entire traced component is treated as an
ancilla value used internally used to perform the larger program but
leaving no memory footprint.
